A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-12-25 paper/paper.tex
paper: misc

\subsection*{Reynolds' parametricity and its unary and binary variants}
In the context of functional programming, Reynolds parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, families over a product of sets, or \emph{correspondences}, can be used in place of relations. Parametricity can then be iterated, and relying on the fibered presentation of correspondences as spans, it has been noted that iterated Reynolds parametricity has the structure of a cubical set~\citep{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds \emph{binary} parametricity by using predicates or families instead of relations or correspondences, and this is a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch.}. This suggests that the definition of augmented semi-simplicial sets and semi-cubical sets can both be seen as instances of a more general construction, which we call $\nu$-sets, of presheaves over a $\nu$-semi-shape category made of words of some cardinal $\nu+1$, where $\nu=1$ gives augmented semi-simplicial sets and $\nu=2$ gives semi-cubical sets.

@@

By equipping the universe construction with a structure of equivalences, as suggested along the lines of \cite{altenkirch15}, we also suspect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\citep{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also definitively ground cubical type theory in iterated parametricity.

>>>>

\subsection*{Reynolds parametricity and its unary and binary variants}
In the context of functional programming, Reynolds parametricity~(\citeyear{reynolds83}) interprets types as relations characterizing the observational behavior of programs of this type. More generally, families over a product of sets, or \emph{correspondences}, can be used in place of relations. Parametricity can then be iterated, and relying on the fibered presentation of correspondences as spans, it has been noted that iterated Reynolds parametricity has the structure of a cubical set~\citep{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds \emph{binary} parametricity by using predicates or families instead of relations or correspondences, and this is a form of realizability~\citep{bernardy12,lasson12,moulin16}. It has then been noted that iterated unary parametricity has the structure of an augmented simplicial set\footnote{Private communication with Hugo Moeneclaey and Thorsten Altenkirch.}. This suggests that the definition of augmented semi-simplicial sets and semi-cubical sets can both be seen as instances of a more general construction, which we call $\nu$-sets, of presheaves over a $\nu$-semi-shape category made of words of some cardinal $\nu+1$, where $\nu=1$ gives augmented semi-simplicial sets and $\nu=2$ gives semi-cubical sets.

@@

By equipping the universe construction with a structure of equivalences, as suggested along the lines of \cite{altenkirch15}, we also suspect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\citep{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also firmly ground cubical type theory in iterated parametricity.
